---
title: "Social Network Analysis"
author: "Michaela Fricova"
date: "2023-03-28"
output: html_document
---

## #5 Community Structure in Zachary Karate Club
This project is the 5/8 part of my mini-series on social network analysis. The goal of this series is to start with the very basics of network analysis, such as with the concepts of centrality and assortativity, and progress towards more advanced topics, including Erdős-Rényi and Configuration models and Exponential family random graphs. In this fifth episode, I investigate the concepts of **community structure** and **Blockmodeling**. To do so, I will use a well known **Zachary Karate Club dataset**. This dataset was first published in 1977 by Wayne W. Zachary and later became a top social network dataset for analysing communities. Zachary collected this dataset over a period of multiple years by continuously tracking friendships between 34 members of a karate club at a US university. During the investigated time period, a dispute arose about raising of the fees which lead to the partitioning of the club into 2 parts. One club part became governed by the President, the other by the Coach. I will now analyze **what led to the splitting of the Zachary karate network**. And, more specifically, **whether community structure affected the network polarization**.

As always, I start by loading the necessary libraries. 

```{r setup, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE}
library(plot.matrix)
library(igraph)
library(igraphdata)
library(intergraph)
#library(ape) 
library(Matrix) 
require(sna)
library(plyr)
library(dbplyr)
library(ggplot2)
```

## Prepping the Data

I use the Zachary's karate network dataset that is built-in in the igraphdata R package. The dataset contains an adjacency matrix, with names of people in the club (vertex attribute denoted name) and their eventual coach after the club was spit into two (vertex attribute denoted Faction). The name Mr. Hi is a pseudonym for the karate instructor, whereas John A. is a pseudonym for the karate club president. The ‘Faction’ vertex attribute gives the faction memberships of the actors. After the split of the club, club members chose their new clubs based on their factions, except actor no. 9, who was in John A.’s faction but chose Mr. Hi’s club. The vertex attribute color codes the same information, albeit this was probably meant for graphing. There are also edge attributes - edge weights in the dataset. which correspond to the number of common activities that the club members took part in.  These activities were: 
1. Association in and between academic classes at the university.
2. Membership in Mr. Hi’s private karate studio on the east side of the city where Mr. Hi taught nights as a part-time instructor.
3. Membership in Mr. Hi’s private karate studio on the east side of the city, where many of his supporters worked out on weekends.
4. Student teaching at the east-side karate studio referred to in (2). This is different from (2) in that student teachers interacted with each other, but were prohibited from interacting with their students.
5. Interaction at the university rathskeller, located in the same basement as the karate club’s workout area.
6. Interaction at a student-oriented bar located across the street from the university campus.
7. Attendance at open karate tournaments held through the area at private karate studios.
8. Attendance at intercollegiate karate tournaments held at local universities. Since both open
and intercollegiate tournaments were held on Saturdays, attendance at both was impossible

```{r, error=FALSE, warning=FALSE, message=FALSE}
data("karate")
summary(karate)
head(karate)
```
```{r}
node_list <- get.data.frame(karate, what = "vertices")
edge_list <- get.data.frame(karate, what = "edges") 
all_nodes <- sort(node_list$name)
plot_data <- edge_list %>% mutate(
        to = factor(to, levels = all_nodes),
        from = factor(from, levels = all_nodes))
ggplot(plot_data, aes(x = from, y = to)) +
      geom_raster() +
      theme_bw() +
      # Because we need the x and y axis to display every node,
      # not just the nodes that have connections to each other,
      # make sure that ggplot does not drop unused factor levels
      scale_x_discrete(drop = FALSE) +
      scale_y_discrete(drop = FALSE) +
      theme(
        # Rotate the x-axis lables so they are legible
        axis.text.x = element_text(angle = 270, hjust = 0),
        # Force the plot into a square aspect ratio
        aspect.ratio = 1,
        # Hide the legend (optional)
        legend.position = "none")
```

```{r pressure, echo=FALSE}
knet <- asNetwork(karate) 
blockmodel(knet,ec = knet %v% "Faction")
kbm <- blockmodel(knet,ec = knet %v% "Faction")$block.model 
g_kar <- sample_sbm(34, pref.matrix = kbm, block.sizes = c(16,18), directed = FALSE)
assortativity(g_kar, c(rep(1,16), rep(2,18))) 
par(mfrow=c(1,2))
plot(g_kar,
     vertex.color = c(rep(1,16), rep(2,18)), 
     vertex.label = NA,
     main = "Fabricated Network")
plot(karate, 
     vertex.label = NA,
     main = "Real Network")
```

```{r}
par(mfrow=c(1,2))
degree_cent_fabricated <- degree(g_kar)
hist(degree_cent_fabricated, 
     col = "#CD7F32",
     main = "Fabricated Network")
degree_cent_real <- degree(karate)
hist(degree_cent_real, 
     col = "#CD7F32",
     main = "Real Network")
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
