---
title: "Social Network Analysis"
author: "Michaela Fricova"
date: "2023-03-28"
output: html_document
---

## #5 Community Structure in Zachary Karate Club
This project is the 5/8 part of my mini-series on social network analysis. The goal of this series is to start with the very basics of network analysis, such as with the concepts of centrality and assortativity, and progress towards more advanced topics, including Erdős-Rényi and Configuration models and Exponential family random graphs. In this fifth episode, I investigate the concepts of **community structure** and **Blockmodeling**. To do so, I will use a well known **Zachary Karate Club dataset**. This dataset was first published in 1977 by Wayne W. Zachary and later became a top social network dataset for analysing communities. Zachary collected this dataset over a period of multiple years by continuously tracking friendships between 34 members of a karate club at a US university. During the investigated time period, a dispute arose about raising of the fees which lead to the partitioning of the club into 2 parts. One club part became governed by the President, the other by the Coach. I will now analyze **what led to the splitting of the Zachary karate network**. And, more specifically, **whether community structure affected the network polarization**.

As always, I start by loading the necessary libraries. 

```{r setup, echo = FALSE, error=FALSE, warning=FALSE, message=FALSE}
library(plot.matrix)
library(igraph)
library(igraphdata)
library(intergraph)
#library(ape) 
library(Matrix) 
require(sna)
library(plyr)
library(dbplyr)
library(ggplot2)
```

## Prepping the Data

I use the Zachary's karate network dataset that is built-in in the igraphdata R package. The dataset contains an adjacency matrix, with names of people in the club (vertex attribute denoted name) and their eventual coach after the club was spit into two (vertex attribute denoted Faction). The name Mr. Hi is a pseudonym for the karate instructor, whereas John A. is a pseudonym for the karate club president. The ‘Faction’ vertex attribute gives the faction memberships of the actors. After the split of the club, club members chose their new clubs based on their factions, except actor no. 9, who was in John A.’s faction but chose Mr. Hi’s club. The vertex attribute color codes the same information, albeit this was probably meant for graphing. There are also edge attributes - edge weights in the dataset. which correspond to the number of common activities that the club members took part in.  These activities were: 
1. Association in and between academic classes at the university.
2. Membership in Mr. Hi’s private karate studio on the east side of the city where Mr. Hi taught nights as a part-time instructor.
3. Membership in Mr. Hi’s private karate studio on the east side of the city, where many of his supporters worked out on weekends.
4. Student teaching at the east-side karate studio referred to in (2). This is different from (2) in that student teachers interacted with each other, but were prohibited from interacting with their students.
5. Interaction at the university rathskeller, located in the same basement as the karate club’s workout area.
6. Interaction at a student-oriented bar located across the street from the university campus.
7. Attendance at open karate tournaments held through the area at private karate studios.
8. Attendance at intercollegiate karate tournaments held at local universities. Since both open
and intercollegiate tournaments were held on Saturdays, attendance at both was impossible

```{r, error=FALSE, warning=FALSE, message=FALSE}
data("karate")
summary(karate)
head(karate)
```

```{r pressure, echo=FALSE}
knet <- asNetwork(karate) 
blockmodel(knet,ec = knet %v% "Faction")
kbm <- blockmodel(knet,ec = knet %v% "Faction")$block.model 
g_kar <- sample_sbm(34, pref.matrix = kbm, block.sizes = c(16,18), directed = FALSE)
assortativity(g_kar, c(rep(1,16), rep(2,18))) 
par(mfrow=c(1,2))
plot(g_kar,
     vertex.color = c(rep(1,16), rep(2,18)), 
     vertex.label = NA,
     main = "Fabricated Network")
plot(karate, 
     vertex.label = NA,
     main = "Real Network")

# Degree centrality 
par(mfrow=c(1,2))
degree_cent_fabricated <- igraph::degree(g_kar, mode = "all")
hist(degree_cent_fabricated, 
     col = '#fb9ca4',
     main = "Fabricated Network",
     xlab = "Degree centrality in the fabricated network",
     breaks = 10,
     xlim=c(0,20))
degree_cent_real <- igraph::degree(karate, mode = "all")
hist(degree_cent_real, 
     col = '#a6d0c8',
     main = "Real Network",
     xlab = "Degree centrality in the real network",
     breaks = 20,
     xlim=c(0,20))
```
# Degree centrality in the real and the fabrincated networks
The plots and adjacency matrices compare the actual karate network to a fabricated network that has similar characteristics and equal block memberships. If we fully understood the data generating process (DGP) of the actual karate network, then we would be able to set up the fabricated network such that it would be indistinguishable from the real network. But, in our case, we observe that the fabricated network does not have exactly the same characteristics as the actual karate network. Primarily, the real network has more vertices with very low degree centrality (i.e. 0-4 connections) and also some vertices with very high degree centrality (i.e. 15+ connections), compared to the fabricated network. Such network structure with a few very popular, 'celebrity' nodes has also been commonly reported in other social networks (e.g. school friendship groups) and it has sometimes been associated with the 'Matthew effect' (Merton, 1968). But in our specific example, the 2 'celebrity' nodes correspond to the fact that there are 2 instructors (1 instructor Mr. Hi and 1 administrator John A) in the Zachary karate club and vast majority of trainees are connected to at least 1 of them. Specifically, instructor Mr. Hi has 16 connections and club administrator John A has 17 connections. In other words, there is a preferential attachment to coaches in the Zachary karate network. In contrast to this preferential attachment structure, our fabricated network is set such that every node within 1 block is stochastically equivalent (Yan et al.2014). That means that the probability of an edge between any two nodes in the fabricated network depends only on which blocks they belong to; connections between all vertices within a block are equally likely. And so, in order to make the fabricated network more similar to the actual karate network, we could use degree-corrected stochastic block modelling (Yan et al.2014). This would relax the assumption of stochastic equivalence within blocks, and instead allow for preferential attachment in the fabricated network. 

```{r}
my_block_sizes <- c(20,20,10)

# Assortative Graph
assortative_mat <- cbind(c(0.3,0.005,0.005),
                         c(0.005,0.3,0.005),
                         c(0.005,0.005,0.3))
assortative_graph <- sample_sbm(50, pref.matrix = assortative_mat, block.sizes=my_block_sizes, directed = FALSE, loops = FALSE)
ablock_layout <- layout_components(assortative_graph)
as_adj(assortative_graph)
par(mfrow=c(1,1))
plot(assortative_graph,
     main = "'Assortative'",
     vertex.color =c(rep(1,20), rep(2,20), rep(3,10)),
     vertex.label=NA,
     vertex.size=8,
     edge.width=0.5,
     layout=ablock_layout)
image(assortative_graph[])

# Ordered Graph
ordered_mat <- cbind(c(0.5,0.3,0.001),
                     c(0.3,0.5,0.3),
                     c(0.001,0.3,0.5))
ordered_graph <- sample_sbm(50, pref.matrix = ordered_mat, block.sizes = my_block_sizes, directed = FALSE, loops = FALSE)
oblock_layout <- layout_components(ordered_graph)
as_adj(ordered_graph)
par(mfrow=c(1,1))
plot(ordered_graph,
     main = "'Ordered'",
     vertex.color =c(rep(1,20), rep(2,20), rep(3,10)),
     vertex.label=NA,
     vertex.size=8,
     edge.width=0.5,
     layout=oblock_layout)
image(ordered_graph[])

# Core-periphery
coreper_mat <-  cbind(c(0.7,0.2,0.1),
                       c(0.2,0.4,0.1),
                       c(0.1,0.1,0.25))
coreper_graph <- sample_sbm(50, pref.matrix = coreper_mat, block.sizes = my_block_sizes, directed = FALSE, loops = FALSE)
cblock_layout <- layout_components(coreper_graph)
as_adj(coreper_graph)
par(mfrow=c(1,1))
plot(ordered_graph,
     main = "'Core-Periphery'",
     vertex.color =c(rep(1,20), rep(2,20), rep(3,10)),
     vertex.label=NA,
     vertex.size=8,
     edge.width=0.5,
     layout=cblock_layout)
image(coreper_graph[])

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
